---
title: "Assignment 1: Manipulation of mammalian sleep data"
author: "Your name and student ID"
date: "Today's date"
output: pdf_document
---
```{r setup, include = FALSE}
source("setup/hw01.RAGS.R")
```

### Instructions 
* Due date: Tuesday, Jan 28 at 10:00pm.
* Late penalty: 50% late penalty if submitted within 24 hours of due date, no 
marks for assignments submitted thereafter.
* This assignment is marked out of 21 points. Points are indicated for each 
question. There are 16 questions total.
* Submission process: Follow the submission instructions on the final page. Make sure you do not remove any `\newpage` tags or rename this file, as this will break the submission.

Helpful hints:

- Every function you need to use was taught during lecture! So you may need to 
revisit the lecture code to help you along by opening the relevant files on data 
hub. Alternatively, you may wish to view the code in the condensed PDFs posted 
in the lectures file on Datahub (Home > ph142-sp20 > lec). Good luck!

- Knit your file early and often to minimize knitting errors! If you copy and 
paste code for the slides, you are bound to get an error that is hard to 
diagnose. Typing out the code is the way to smooth knitting! We recommend 
knitting your file each time after you write a few sentences/add a new code 
chunk, so you can detect the source of knitting errors more easily. This will 
save you and the GSIs from frustration!

- If your code runs off the page of the knitted PDF then you will LOSE POINTS! To
avoid this, have a look at your knitted PDF and ensure all the code fits in the 
file. When it doesn't, go back to your .Rmd file and add spaces (new lines) using
the return or enter key so that the code runs onto the next line.
 
--------------------------------------------------------------------------------
\newpage

Begin by knitting this document by pushing the "Knit" button above. As you fill 
in code and text in the document, you can re-knit (push the button again) and 
see how the document changes. It is important to re-knit often, because if there
is any error in your code, the file will not generate a PDF, so our advice is to
knit early and often!

### Using `dplyr` to investigate sleep times in mammals

The data file `sleep.csv` contains the sleeptimes and weights for a set of mammals. 
Hit the green arrow icon in the line below to execute the two lines of code in 
the code chunk, or execute them line by line by placing your cursor on the first 
line and hitting cmd + enter on Mac or ctrl + enter on PC. 

```{r import-and-save-data}
library(dplyr)
library(readr)
sleep <- read_csv("sleep.csv")
```

- The `library` command loads the library `dplyr` into memory. 
- The `readr` library contains functions to read in the dataset.
- The `dplyr` library contains functions we will use to manipulate data.

Notice that an object called `sleep` appeared in the Environment tab under "Data". 

\newpage

**1. [2 points] Use four useful functions discussed in lecture to examine the sleep data set:**

```{r four-functions}
# Text inside a code chunk that begins with "#" is called a comment. 
# We sometimes use comments to explain code to you in plain English.
# Write your four functions below these comments, replacing the placeholder
# text "<<<<YOUR CODE HERE>>>>". Remember, code does *not* begin with a "#"

"<<<<YOUR CODE HERE>>>>"
"<<<<YOUR CODE HERE>>>>"
"<<<<YOUR CODE HERE>>>>"
"<<<<YOUR CODE HERE>>>>"


# Then, assign p1 to a vector of your function names, in alphabetical order.
# For example, assigning p0 to a vector of fruits looks like this:
# p0 <- c("apple", "banana", "orange")

p1 <- "<<<<YOUR CODE HERE>>>>"

check_problem1()
```

\newpage

Description of the variables found in the sleep dataset:

| Column name	| Description |
|-------------|-------------|
|name	|common name|
|genus	|taxonomic rank|
|vore	|carnivore, omnivore or herbivore?|
|order	| taxonomic rank|
|conservation	|the conservation status of the mammal|
|sleep_total|	total amount of sleep, in hours|
|sleep_rem |	Rapid eye movement (REM) sleep, in hours |
|sleep_cycle|	length of sleep cycle, in hours|
|awake|	amount of time spent awake, in hours|
|brainwt|	brain weight in kilograms|
|bodywt|	body weight in kilograms|

**2. [2 points] Write code to select a set of columns. Specifically select the `awake`, `brainwt`, and `bodywt` columns. Assign this smaller dataset to a data frame called `sleep_small`**

```{r select-code}
sleep_small <- "<<<<YOUR CODE HERE>>>>"

check_problem2()
```
\newpage

**3. [1 point] To select a range of columns by name, use the “:” (colon) operator. Redo the selection for question 1, but use the colon operator. Assign this to `sleep_small_colon`. Note that this returns the same data frame as the previous problem, but is not recommended in practice because it depends on the ordering of the columns and isn't explicit in the columns that are selected, whereas selection by name offers much higher readability for someone else looking at your code later on.**

```{r select-code-partII}
sleep_small_colon <- "<<<<YOUR CODE HERE>>>>"
  
check_problem3()
```
\newpage

**4. [1 point] Select all the columns except for the `vore` variable. Assign this to `sleep_no_vore`.**

```{r select-code-partIII}
sleep_no_vore <- "<<<<YOUR CODE HERE>>>>"

check_problem4()
```
\newpage

**5. [1 point] Run the following chunk of code.**

```{r select-code-partIV}
select(sleep, starts_with("sl"))
```

**What does it return? Uncomment one of the possible choices.**
```{r histogram-GDP-2006}
# p5 <- "returns the number of columns that start with sl"
# p5 <- "returns all columns that start with sl"
# p5 <- "returns all rows that start with sl"
# p5 <- "returns all animals whose names start with sl""

check_problem5()
```

It returns the columns that start with sl: `sleep_total`, `sleep_rem`, `sleep_cycle`

\newpage

**6. [1 point] Rewrite the previous chunk of code using the pipe operator. Assign this to `sleep_sl`.**

```{r use-the-pipe}
sleep_sl <- "<<<<YOUR CODE HERE>>>>"

check_problem6()
```
\newpage

**7. [1 point] Filter the rows for mammals that sleep a total of more than 16 hours. Assign this to `sleep_over16`.**

```{r filter-code}
sleep_over16 <- "<<<<YOUR CODE HERE>>>>"

check_problem7()
```
\newpage

**8. [2 points] Filter the rows for mammals that sleep a total of more than 16 hours and have a body weight of greater than 1 kilogram. Assign this to `sleep_mammals`.**

```{r filter-code-again}
sleep_mammals <- "<<<<YOUR CODE HERE>>>>"

check_problem8()
```
\newpage

**9. [1 point] Suppose you are specifically interested in the sleep of horses and giraffes. Assign `sleep_hg` to a data frame for horses and giraffes only. **

```{r horses-and-giraffes}
sleep_hg <- "<<<<YOUR CODE HERE>>>>"

check_problem9()
```
\newpage

**10. [1 point] Order the dataset by sleep time from shortest sleep time to longest sleep time. Assign this to `sleep_time`.**

```{r order-by-time}
sleep_time <- "<<<<YOUR CODE HERE>>>>"

check_problem10()
```
\newpage

**11. [1 point] Now order for longest sleep time to shortest sleep time. Assign this to `sleep_time_rev`.**

```{r flip-order}
sleep_time_rev <- "<<<<YOUR CODE HERE>>>>"

check_problem11()
```
\newpage

**12. [2 points] Suppose you are interested in the order of sleep time, but according to whether the animal is a carnivore, herbivore, or omnivore. Rewrite the above statement to order sleep time according to the type of "-vore" that then animal is:**

```{r order-within}
sleep_time_vore <- "<<<<YOUR CODE HERE>>>>"

check_problem12()
```
\newpage

**13. [1 point] Create a new column called `rem_proportion` which is the ratio of rem sleep to total amount of sleep. Assign this new data frame to `sleep_ratio`.**

```{r new-variable}
sleep_ratio <- "<<<<YOUR CODE HERE>>>>"

check_problem13()
```
\newpage

**14. [1 point] Add a second column called `bodywt_grams` which is the bodywt column in grams.**

```{r another-new-variable}
sleep_r_bw <- "<<<<YOUR CODE HERE>>>>"

check_problem14()
```
\newpage

**15. [1 point] Calculate the average sleep time across all the animals in the dataset using a `dplyr` function and assign it to the variable `avg_sleep_time`. Your answer should be a data frame of 1 observation and 1 variable called `sleep_avg`**

```{r calc-average}
avg_sleep_time <- "<<<<YOUR CODE HERE>>>>"

check_problem15()
```
\newpage

**16. [2 points] Calculate the average sleep time for each type of "-vore". Hint: you'll need to use two `dplyr` functions! The column names should be `vore` and `sleep_avg`.**

```{r calc-average-within}
avg_by_vore <- "<<<<YOUR CODE HERE>>>>"

check_problem16()
```

\newpage

### Check your score
```{r check-total-score}
# Just run this chunk.
total_score()
```

\newpage

### Submission

For assignments in this class, you'll be submitting using the **Terminal** tab in the pane below. In order for the submission to work properly, make sure that:

1. Any image files you add that are needed to knit the file are in the `src` folder and file paths are specified accordingly. 
2. You **have not changed the file name** of the assignment.
3. The file knits properly.

Once you have checked these items, you can proceed to submit your assignment.

1. Click on the **Terminal** tab in the pane below.
2. Copy-paste the following line of code into the terminal and press enter.

cd; cd ph142-sp20/hw/hw01; python3 turn_in.py

3. Follow the prompts to enter your Gradescope username and password.
4. If the submission is successful, you should see "Submission successful!" appear as output.
5. If the submission fails, try to diagnose the issue using the error messages--if you have problems, post on Piazza under the post "Submission Issues". 

We will not be accepting late submissions **no matter the reason**, including submission issues, so be sure to submit early enough to have time to diagnose issues if problems arise.
