---
title: 'Lab 10: Proportions and Inference'
output: pdf_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include = FALSE}
# Don't change these lines, just run them!
source("setup/lab10.RAGS.R")
```

### Tests of changes in sex ratios based on a single sample

There is a long literature studying changes in sex-ratios of births due to 
stressful events, such as 9/11. In today's lab, we consider a relatively small 
study that recorded biomarkers of stress on pregnancy.  In the group of subjects
that had the highest markers of stress (based on cortisol), there were 14 births
to males out of a total of 38.

In this lab, we will compare the four methods we learned to calculate CIs for
proportions. Recall that two of these methods involved hand calculations (though
we can treat R as if it were a calculator) and two of the methods used built-in
R functions.

1. Use the Normal approximation to construct a 95% (using z = 1.96) confidence interval
in this high stress group. Store your answer to p1 using the following format:

```{r eval = F}
lowerbound <- 0
upperbound <- 1

example <- c(lowerbound, upperbound)
```

We also called this specific method of 
constructing the CI the "large sample method"

```{r large-sample-method}
p1 <- "You answer here"
p1

check_problem1()
```



2. Create the 95% CI again, this time using the R function that implements the Wilson Score method with a continuity correction. 

```{r wilson-score}
p2 <- "your answer here"
p2


check_problem2()
```



3.  Create the 95% CI (using z = 1.96) again, this time using the Plus 4 method. Store your answer to p3 using the following format.

```{r eval = F}
example <- c(lowerbound, upperbound)
```


```{r plus-four}
p3 <- "your answer here"
p3



check_problem3()

```




4.  Create the 95% CI again, this time using the R function that implements the Clopper Pearson (Exact) method.

```{r exact-method}
p4 <- "your answer here"
p4


check_problem4()

```



5. Summarize the four methods' estimates in the following table. Do they include 
the null hypothesized value for the sex ratio?

| Method        | 95% Confidence Interval | 
|---------------|:-----------------------:|
| Large sample  | AA.A% to AA.A%          |
| Wilson Score* | AA.A% to AA.A%          | 
| Plus four     | AA.A% to AA.A%          | 
| Exact         | AA.A% to AA.A%          | 

* with continuity correction.

6. Here is a code template to help you to graphically present these estimates. 
Graphical presentations of estimates and their CIs is very useful for assessing
whether the CIs overlap the null hypothesized value and tends to be better than
presenting tables of estimates to readers of your research.

```{r}
# First make a tibble (an easy way to make a data frame) with the data about
# each confidence interval. To do this, replace each instance of 0.00 with the
# estimate from your calculations above.

library(ggplot2)
library(tibble)
sex_CIs <- tibble(method   = c("large sample", "Exact", "Wilson", "Plus 4"),
                  lower_CI = c(0.0           , 0.0    ,  0.0    ,  0.0),
                  upper_CI = c(0.0           , 0.0    ,  0.0    ,  0.0),
                  estimate = c(0.0           , 0.0    ,  0.0    ,  0.0)
                  )

# Build the ggplot incrementally, to understand how it works.
# Step 1: (qu: why do we put a horizontal line at 50?)
ggplot(data = sex_CIs, aes(x = method, y = estimate)) + 
  geom_point() + 
  geom_hline(aes(yintercept = 50), lty = 2)

# Step 2:
ggplot(data = sex_CIs, aes(x = method, y = estimate)) + 
  geom_point() +
  geom_hline(aes(yintercept = 50), lty = 2) +
  geom_segment(aes(xend = method, y = lower_CI, yend = upper_CI)) +
  labs(y = "Estimate with 95% CI")


```

What does `geom_segment()` do? In particular, what do `x`, `xend`, `y` and `yend`
specify in this case?



7. Based on this plot, what can you say about the confidence intervals for the 
sex ratio in the high stress group?



8. If you have time, repeat the above analysis for the group with low stress. There
were 25 births to this group, of which 17 of them were to males.

```{r}
# your code here

```

9. If you recreated the graph for the low stress group, what can you say about 
the confidence intervals for the sex ratio in this group?





\newpage

### Check your score

Click on the middle icon on the top right of this code chunk (with the downwards gray arrow and green bar) to run all your code in order. Then, run this chunk to check your score.

```{r check-total-score}
# Just run this chunk.
total_score()
```

\newpage

### Submission

For assignments in this class, you'll be submitting using the **Terminal** tab in the pane below. In order for the submission to work properly, make sure that:

1. Any image files you add that are needed to knit the file are in the `src` folder and file paths are specified accordingly. 
2. You **have not changed the file name** of the assignment.
3. The file is saved (the file name in the tab should be **black**, not red with an asterisk).
4. The file knits properly.

Once you have checked these items, you can proceed to submit your assignment.

1. Click on the **Terminal** tab in the pane below.
2. Copy-paste the following line of code into the terminal and press enter.

cd; cd ph142-sp20/lab/lab10; python3 turn_in.py

3. Follow the prompts to enter your Gradescope username and password. When entering your password, you won't see anything come up on the screen--don't worry! This is just for security purposes--just keep typing and hit enter.
4. If the submission is successful, you should see "Submission successful!" appear as output.
5. If the submission fails, try to diagnose the issue using the error messages--if you have problems, post on Piazza. 

The late policy will be strictly enforced, **no matter the reason**, including submission issues, so be sure to submit early enough to have time to diagnose issues if problems arise.
