---
title: "Assignment 8"
author: "Your name and student ID"
date: "Today's date"
output:
  pdf_document: default
---

```{r, warning=F, message=F, echo=F}
library(dplyr)
library(ggplot2)
```

* Due date: Tuesday, April 14 10:00pm.
* Late penalty: 50% late penalty if submitted within 24 hours of due date, no 
marks for assignments submitted thereafter.
* This assignment is marked out of 20. Marks are indicated for each 
question. 
* Remember: Autograder is meant as sanity check ONLY. It will not tell you if you have the correct answer. It will tell you if you are in the ball park of the answer so *CHECK YOUR WORK*
* Submission process: Follow the submission instructions on the final page. Make sure you do not remove any `\newpage` tags or rename this file, as this will break the submission.

```{r setup, include = FALSE}
# Don't change these lines, just run them!
source("setup/hw08.RAGS.R")
```

Helpful hints:

-  Every function you need to use was taught during lecture! So you may need to revisit the lecture code to help you along by opening the relevant files on Datahub. Alternatively, you may wish to view the code in the condensed PDFs posted on the course website. Good luck!

- Knit your file early and often to minimize knitting errors! If you copy and paste code for the slides, you are bound to get an error that is hard to diagnose. Typing out the code is the way to smooth knitting! We recommend 
knitting your file each time after you write a few sentences/add a new code 
chunk, so you can detect the source of knitting errors more easily. This will save you and the GSIs from frustration! **You must knit right before submitting**

- If your code runs off the page of the knitted PDF then you will LOSE POINTS! To avoid this, have a look at your knitted PDF and ensure all the code fits in the file. When it doesn't, go back to your .Rmd file and add spaces (new lines) using the return or enter key so that the code runs onto the next line.

\newpage

Round your answers to 3 decimal places in general.

--------------------------------------------------------------------------------

In two wards for elderly patients in a local hospital the following 
levels of hemoglobin (grams per liter) were found for a simple random sample of patients from each ward.:

Ward A: 
```{r}
ward_a <- c(12.2, 11.1, 14.0, 11.3, 10.8, 12.5, 12.2, 11.9, 13.6, 12.7, 13.4, 13.7)
```

Ward B: 
```{r}
ward_b <- c(11.9, 10.7, 12.3, 13.9, 11.1, 11.2, 13.3, 11.4, 12.0, 11.1)
```

1. [1 point] In one ggplot, create two box plots to compare the hemoglobin values for Ward A and Ward B. Also plot the raw data as points, overlaid on top of the box plots. 

```{r make-data-frame}

hemoglobin <- data.frame(hemo_level = c(ward_a, ward_b), 
                         ward = c(rep("Ward A", 12), rep("Ward B", 10)))
```

```{r make-box-plot}

p1 <- "YOUR ANSWER HERE"
p1
check_problem1()
```

\newpage

2. [1 points] Comment on the similarities/differences portrayed by the plots, keeping in mind that the sample size is relatively smallfor these two wards.

[TODO: YOUR ANSWER HERE]


\newpage

3. [2 points] What two assumptions do you need to make to use any of the t-procedures? Because each ward has a rather small sample size (n < 12 for both), what two characteristics of the data would you need to check for to ensure that the t-procedures can be applied?

[TODO: YOUR ANSWER HERE]


\newpage

4. [3 points] Using only `dplyr` and `*t` functions, create a 95% confidence interval for the mean difference between Ward A and Ward B. You can do this by using `dplyr` to calculate the inputs required to calculate the 95% CI, and then plugging these values in on a separate line of code (or using your calculator). Use a degrees of freedom of 19.515 (You don't need to calculate the degrees of freedom, you can use this value directly). Show your work and interpret the mean difference and its 95% CI. Round your solution to 3 decimal places.

```{r}
# YOUR CODE HERE

# THEN, ASSIGN YOUR FINAL ANSWERS BELOW:
CI_lowerbound <- "YOUR ANSWER HERE"
CI_upperbound <- "YOUR ANSWER HERE"

check_problem4()
```

\newpage

5. [1 points] Interpret the mean difference and its 95% CI you just calculated.

[TODO: YOUR ANSWER HERE]


\newpage

Perform a two-sided t-test for the difference between the two samples, where
the null hypothesis is that the underlying means are the same. Start by 
writing down the null and alternate hypotheses, then calculate the test statistic
(showing your work) and p-value. Continue to assume that the degrees of freedom
is 19.515. Verify the p-value by running the t-test using R's built in function.
Show the output from that test. Hint: to perform the t-test using R's built in 
function, you need to pass the function an x and y argument, where x includes 
that values for Ward A and Y includes the values for Ward B. `dplyr`'s `filter()`
and `pull()` functions will be your friends.


6. [1 points] Calculate the t-test statistics

```{r}
t_statistics <- "YOUR ANSWER HERE"
t_statistics


check_problem6()
```


\newpage

7. [1 points] We need to compare this t-statistic to a t distribution with 19.515 degrees of freedom:

```{r}
p_value <- "YOUR ANSWER HERE"
p_value


check_problem7()
```

\newpage

8. [2 points] Interpret the p value you got in the context of the this question. Are there evidence against null hypothesis?

[TODO: YOUR ANSWER HERE]

\newpage

Check this against the `t.test` output:
```{r}
t.test(x = hemoglobin %>% filter(ward == "Ward A") %>% pull(hemo_level), 
       y = hemoglobin %>% filter(ward == "Ward B") %>% pull(hemo_level), 
       alternative = "two.sided")
```

--------------------------------------------------------------------------------
\newpage

The time to perform open heart surgery is normally distributed. Sixteen 
patients (chosen as a simple random sample from a hospital) underwent open heart 
surgery that took the following lengths of time (in minutes).

```{r}
op_time <- c(247.8648, 258.4343, 315.6787, 268.0563, 269.9372, 320.6821, 
             280.5493, 225.3180, 243.8207, 251.5388, 304.9706, 277.3140, 
             278.6247, 269.3418, 248.0131, 322.9812)
surg_data <- data.frame(op_time)
```

9. [1 point] You wish to know if the mean operating time of open heart surgeries at this hospital exceeds four hours. Set up appropriate hypotheses for investigating
this issue.

[TODO: YOUR ANSWER HERE]


\newpage

10. [1 points] Test the hypotheses you formulated in part (a). Report the p-value. (Do not use the `t.test` function for this question)

```{r}
p_value_10 <- "YOUR ANSWER HERE"
p_value_10

check_problem10()
```

\newpage

11. [1 points] What are your conclusions in the context of the question? 

[TODO: YOUR ANSWER HERE]


\newpage

12. [3 points] Construct a 95% CI on the mean operating time (in hours).

```{r}
# YOUR CODE HERE

# THEN, ASSIGN YOUR ANSWERS BELOW:
CI_lowerbound_12 <- "YOUR ANSWER HERE"
CI_upperbound_12 <- "YOUR ANSWER HERE"

check_problem12()
```


\newpage

--------------------------------------------------------------------------------

13. [1 point] Suppose you were testing the hypotheses $H_0: \mu_d = 0$ and $H_a: \mu_d \neq 0$ in a paired design and obtain a p-value of 0.21. Which one of the following could be a possible 95% confidence interval for $\mu_d$?

```{r}
# Uncomment one of the following choices:
# p13 <- "-2.30 to -0.70"
# p13 <- "-1.20 to 0.90"
# p13 <- "1.50 to 3.80"
# p13 <- "4.50 to 6.90"

check_problem13()
```

\newpage

--------------------------------------------------------------------------------

14. [1 point] Suppose you were testing the hypotheses $H_0: \mu_d = 0$ and $H_a: \mu_d \neq 0$ in a paired design and obtain a p-value of 0.02. Also suppose you computed confidence intervals for $\mu_d$. Based on the p-value which one of the following is true?

```{r}
# Uncomment one of the following choices:
# p14 <- "Both a 95% CI and a 99% CI will contain 0."
# p14 <- "A 95% CI will contain 0, but a 99% CI will not."
# p14 <- "A 95% CI will not contain 0, but a 99% CI will."
# p14 <- "Neither a 95% CI nor a 99% CI interval will contain 0." 

check_problem14()
```

\newpage

### Check your score

Click on the middle icon on the top right of this code chunk (with the downwards gray arrow and green bar) to run all your code in order. Then, run this chunk to check your score.
```{r check-total-score}
# Just run this chunk.
total_score()
```

\newpage

### Submission

For assignments in this class, you'll be submitting using the **Terminal** tab in the pane below. In order for the submission to work properly, make sure that:

1. Any image files you add that are needed to knit the file are in the `src` folder and file paths are specified accordingly. 
2. You **have not changed the file name** of the assignment.
3. The file is saved (the file name in the tab should be **black**, not red with an asterisk).
4. The file knits properly.

Once you have checked these items, you can proceed to submit your assignment.

1. Click on the **Terminal** tab in the pane below.
2. Copy-paste the following line of code into the terminal and press enter.

cd; cd ph142-sp20/hw/hw08; python3 turn_in.py

3. Follow the prompts to enter your Gradescope username and password. When entering your password, you won't see anything come up on the screen--don't worry! This is just for security purposes--just keep typing and hit enter.
4. If the submission is successful, you should see "Submission successful!" appear as output.
5. If the submission fails, try to diagnose the issue using the error messages--if you have problems, post on Piazza. 

The late policy will be strictly enforced, **no matter the reason**, including submission issues, so be sure to submit early enough to have time to diagnose issues if problems arise.


